Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-02-28 paper/paper.tex
pass on intro

This paper focuses on rule-based \nids{}, which is a very popular kind
of NIDS used in industry (see Section~\ref{sec:background}). A
\emph{rule-based intrusion detector}\footnote{\aka\ signature-based
  intrusion detector.} checks if the network traffic matches a fixed
set of rules. Figure~\ref{fig:synflood-example} shows an example rule
of Suricata~\cite{suricata}, a popular open-source \nids{}. This
particular rule prescribes a method to capture a denial-of-service
attack to a server by matching specific conditions about the network
traffic~\cite{understanding-dos}. Relevant properties about the
traffic of interest appear in bold in this rule (see
Section~\ref{sec:suri-metas-coverage}). Deployments of rule-based
\nids\ are restricted to a fixed set of rules defined by the network
system administrator. Updating these rules is tedious and error-prone,
but it is important as attackers constantly create new strategies to
circumvent existing rules. IT-security companies capitalize on this
phenomena and offer rulesets on the
market~\cite{proofpoint-etpro,snort-rule-subscriptions}.

This paper proposes \tname{}\footnote{Abbreviation for
  \textbf{Sy}nthesis of Su\textbf{ri}cata R\textbf{u}le\textbf{s}.}, a
technique that uses machine intelligence to automatically synthesize
rules for rule-based \nids. Our goal is to facilitate the creation
process of rules and, consequently, harden the protection of \nids\ to
network attacks. \tname{} synthesizes rules from malicious and benign
traffic and from examples of correct rules.

@@

Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}.

A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied.
In this example, a
message will be sent to system administrators if the rule pattern is
satisfied. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. For this rule, the header is \CodeIn{tcp \$HOME\_NET any ->
  \$EXTERNAL\_NET any}. It instructs Suricata to inspect \CodeIn{tcp}
traffic flowing from any port in the home network to any other address
outside the home network. The variables \CodeIn{\$HOME\_NET} and
\CodeIn{\$EXTERNAL\_NET} are configurable. The rule options come after
the header. It is a semi-colon-separated sequence of key-value
pairs. Rule options serve to document the analyzed traffic (\eg{},
``msg'', ``classtype'', and ``sid'') and to characterize the attack
pattern (\eg, ``flags'', ``threshold'', and ``content''). \tname{}
produces options of the second kind, which are \emph{relevant} to
capture the attack. For example, the option \CodeIn{content:<arg>} checks if
the argument is present in the payload of the message.

\vspace{-1ex}
\begin{table}[h!]
  \footnotesize
  \caption{\label{table:rules}Example options of Suricata.}
  \vspace{-2ex}
  \centering
  \begin{tabular}{clp{5.5cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & minimum threshold for a rule before it generates alerts\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}
%\vspace{-1ex}

Table~\ref{table:rules} shows a sample of the options of
Suricata. Column ``\#'' shows the option number, column ``Name'' shows
the name of the option, and column ``Description'' presents a short
description of the option purpose. We found that the option
\CodeIn{content} is particularly prevalent in rules from public
databases\MyComment{ (see Figure~\ref{fig:distribution-contents})}. This
option takes a sequence of bytes as argument (\eg{},
\CodeIn{content:''SELECT ''}) and checks if that sequence is present
in the payload of the message. Detailed description of the Suricata
rules can be found elsewhere~\cite{suri-rule-format}. It is worth
noting that Suricata supports a total of \totoptions\ distinct options
covering \numproto\ distinct protocols.

@@

\vspace{-1ex}
\begin{figure}[h!]
  \lstinputlisting[language=C,numbers=none,keywords={dsize,itype}]{pingscan.suricata}
  \vspace{-2ex}  
  \caption{Suricata rule for Ping Scan attack.}
  \label{fig:pingscan-example}
\end{figure}
\vspace{-3ex}

@@

The SYN Flood attack is a denial-of-service attack that exploits a vulnerability in the TCP/IP handshake
to establish a TCP connection~\cite{cloudfare-synflood}. The handshake
works as follows in normal circumstances. First, a client sends a SYN
packet to the server, requesting a connection. Second, the server
responds with a SYN-ACK packet to the client. Third, the client
responds with an ACK message and the connection is established. Aware
of the protocol, an attacker sends multiple SYN packets to different
ports of a server, often using fake IP addresses. Then, after the
server responds with a SYN-ACK packet, the client keeps sending other
SYN packets to avoid the connection to time out. Without proper
protection, the server accepts these malicious requests and eventually
legitimate requests cannot be satisfied due to resource exhaustion.
Figure~\ref{fig:synflood-example} shows a Suricata rule that captures
this attack. This rule was obtained from the Emerging Threats Open
Ruleset~\cite{emerging-threats-open}. The relevant options in the
rule appear in bold---the option \CodeIn{flags: S,12}, identifies a
SYN packet in a TCP packet and the option \CodeIn{threshold: type
  both, track by\_dst, count 5000, seconds 5} indicates that a high
volume of such packets should be requested in a short period of
time. The other options are for documentation purposes.

>>>>

This paper focuses on rule-based \nids{}, which are very popular in
industry~\cite{proofpoint-etpro,snort-rule-subscriptions}. A
\emph{rule-based intrusion detector}\footnote{\aka\ signature-based
  intrusion detector.} checks if the network traffic matches a fixed
set of rules. Figure~\ref{fig:pingscan-example} shows an example rule
of Suricata~\cite{suricata}, a popular open-source \nids{}. This
particular rule prescribes a method to capture an Active
Reconnaissance attack for discovering which ports of a publicly
accessible machine are open for connection. Rulesets exist for
Snort~\cite{snort} and Suricata~\cite{suricata}, providing protection
against existing attacks. Unfortunately, attackers actively look for
new mechanisms to exploit systems. For example, Figure~\Fix{plot}
shows the increase in number of rules added to the ``Emerging Threats
Open Ruleset''---a popular public data set of \suri\ rules---in the
past 12 months.

\Fix{add figure here}

Intuitively, rulesets need to be constantly revised to protect a
system against attackers. Unfortunately, creating these rules is
time-consuming as the language for describing rules is non-trivial
(see Section~\ref{sec:example-suricata-rules}) and it is error-prone
as checking whether the rule captures only malicious traffic is
non-trivial.

%% IT-security companies capitalize on this phenomena and
%% offer rulesets on the
%% market.

This paper proposes \tname{}\footnote{Abbreviation for
  \textbf{Sy}nthesis of Su\textbf{ri}cata R\textbf{u}le\textbf{s}.}, a
technique that uses machine intelligence to automatically synthesize
rules for rule-based \nids. Our goal is to facilitate the creation of
rules and, consequently, harden the protection of \nids\ to network
attacks. \tname{} synthesizes rules from malicious and benign traffic
and from examples of correct rules.

\Fix{
denial-of-service
attack to a server by matching specific conditions about the network
traffic~\cite{understanding-dos}. Relevant properties about the
traffic of interest appear in bold in this rule (see
Section~\ref{sec:suri-metas-coverage}). Deployments of rule-based
\nids\ are restricted to a fixed set of rules defined by the network
system administrator.}\Mar{check...}

@@

Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}.
A Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied.  In this example, a
message will be sent to system administrators if the rule pattern is
satisfied. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. The header of this rule is \CodeIn{icmp any any -> any
  any}. It instructs Suricata to inspect \CodeIn{icmp} traffic flowing
from any machine and port to any machine and port.\MyComment{ The variables
\CodeIn{\$HOME\_NET} and \CodeIn{\$EXTERNAL\_NET} are
configurable.} The rule options come after the header. It is a
semi-colon-separated sequence of key-value pairs. Rule options serve
to document the analyzed traffic (\eg{}, ``msg'', ``classtype'', and
``sid'') and to characterize the attack pattern (\eg, ``flags'',
``threshold'', and ``content''). \tname{} produces options of the
second kind, which are \emph{relevant} to capture the attack. For
example, the option \CodeIn{content:<arg>} checks if the argument is
present in the payload of the message.

\vspace{-1ex}
\begin{table}[h!]
  \footnotesize
  \caption{\label{table:rules}Example of non-documentation options of Suricata.}
  \vspace{-2ex}
  \centering
  \begin{tabular}{clp{5.5cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & minimum threshold for a rule before it generates alerts\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}
%\vspace{-1ex}

Table~\ref{table:rules} shows a sample of non-documentation options of
Suricata. Column ``\#'' shows the option number, column ``Name'' shows
the name of the option, and column ``Description'' summarizes
purpose. The option \CodeIn{content} is very common in rules to
capture http attacks, which are the most common kind of rules. This
option takes a sequence of bytes as argument (\eg{},
\CodeIn{content:''SELECT ''}) and checks if that sequence is present
in the payload of the message. Detailed description of the Suricata
rules can be found elsewhere~\cite{suri-rule-format}. It is worth
noting that Suricata supports a total of \totoptions\ distinct options
covering \numproto\ distinct protocols.

@@

%\vspace{-1ex}
\begin{figure}[h!]
  \lstinputlisting[language=C,numbers=none,keywords={dsize,itype}]{pingscan.suricata}
  \vspace{-2ex}  
  \caption{Suricata rule for Ping Scan attack.}
  \label{fig:pingscan-example}
\end{figure}
\vspace{-3ex}

@@

The SYN Flood attack is a denial-of-service attack that exploits a vulnerability in the TCP/IP handshake
to establish a TCP connection~\cite{cloudfare-synflood}. The handshake
works as follows in normal circumstances. First, a client sends a SYN
packet to the server, requesting a connection. Second, the server
responds with a SYN-ACK packet to the client. Third, the client
responds with an ACK message and the connection is established. Aware
of the protocol, an attacker sends multiple SYN packets to different
ports of a server, often using fake IP addresses. Then, after the
server responds with a SYN-ACK packet, the client keeps sending other
SYN packets to avoid the connection to time out. Without proper
protection, the server accepts these malicious requests and eventually
legitimate requests cannot be satisfied due to resource exhaustion.

Figure~\ref{fig:synflood-example} shows a Suricata rule that captures
this attack. This rule was obtained from the Emerging Threats Open
Ruleset~\cite{emerging-threats-open}. The relevant options in the
rule appear in bold---the option \CodeIn{flags: S,12}, identifies a
SYN packet in a TCP packet and the option \CodeIn{threshold: type
  both, track by\_dst, count 5000, seconds 5} indicates that a high
volume of such packets should be requested in a short period of
time. The other options are for documentation purposes.
